quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-11-30 paper-quollr.Rmd
Move usage to before implementation and re-order many things in implementation

@@

@@

The following demonstration of the package's functionality assumes `quollr` has been loaded. To begin the workflow, user needs two inputs: the high-dimensional dataset and the corresponding NLDR data. The high-dimensional data must contain a unique `ID` column, with data columns prefixed by the letter `"x"` (e.g., `x1`, `x2`, etc.). The NLDR dataset should include embedding coordinates labeled as `emb1` and `emb2`, ensuring one-to-one correspondence with the high-dimensional data through the shared `ID`.

To illustrate the workflow, we use the built-in example dataset `scurve`, a $7\text{-}D$ simulated dataset consisting of $1000$ observations. The first three variables define a $3\text{-}D$ S-shaped manifold, while the remaining four variables introduce low-magnitude uniform noise, yielding a structured yet noisy high-dimensional dataset.

```{r, echo=TRUE}
data("scurve")
```

For this example, we use the UMAP layout, which is used `n_neighbors =` $46$ and `min_dist =` $0.9$ as hyper-parameters.

```{r, echo=TRUE, eval=FALSE}
library(umap)
library(dplyr)

scurve_umap <- umap(
  scurve |> select(-ID),
  config = modifyList(umap.defaults, list(
    n_neighbors = 46,
    n_components = 2,
    min_dist = 0.9
  ))
)$layout |>
  as_tibble(.name_repair = ~ paste0("emb", 1:2)) |>
  mutate(ID = scurve$ID)
```

## Main function

The mains steps for the algorithm can be executed by the main function `fit_highd_model()`, or can be run separately for more flexibility.

This function requires several parameters: the high-dimensional data (`highd_data`), the emdedding data (`nldr_data`), the number of bins along the x-axis (`b1`), the buffer amount as a proportion of data (`q`), and benchmark value to extract high density hexagons (`hd_thresh`). The function returns an object of class `highd_vis_model` containing the scaled NLDR object (`nldr_scaled_obj`) with three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); the hexagonal object (`hb_obj`), the fitted model in both $2\text{-}D$ (`model_2d`), and $p\text{-}D$ (`model_highd`), and triangular mesh (`trimesh_data`).

```{r, echo=TRUE, eval=FALSE}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
```

## Scaling the data

@@

## Computing hexagon grid configuration

The function `calc_bins_y()` determines the configuration of the hexagon grid by computing the number of bins along the y-axis (`b2`), the hexagon binwidth (`a1`), and binheight (`a2`). This function accepts (i) an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); (ii) the number of bins along the x-axis (`b1`), and (iii) the buffer amount as a proportion (`q`). The buffer ensures that the grid fully covers the data space by extending one hexagon width ($a_1$) and height ($a_2$) beyond the observed data in all directions. By default, $q = 0.1$, but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.

@@

## Binning the data

@@

### Mapping the points to their corresponding hexagonal bins

## Obtaining bin centroids

@@

## Indicating neighbors by line segments connecting centroids

To represent the neighborhood structure of hexagonal bins in a $2\text{-}D$ layout, we employ Delaunay triangulation [@lee1980;@albrecht2024] on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

The `tri_bin_centroids()` function generates a triangulation object from the x and y coordinates of hexagon centroids using the `interp::tri.mesh()` function [@albrecht2024]. This triangulation forms the structural basis for identifying adjacent bins.

@@

The `gen_edges()` function uses this triangulation object to extract line segments between neighboring bins. It constructs a unique set of bin-to-bin connections by identifying the triangle edges and filtering duplicate or reversed links. Each edge is then annotated with its start and end coordinates, and a Euclidean distance is computed using the helper function `calc_2d_dist()`.

>>>>

The user needs two inputs: the high-dimensional dataset and the corresponding NLDR data. The high-dimensional data must contain a unique `ID` column, with data columns prefixed by the letter `"x"` (e.g., `x1`, `x2`, etc.). The NLDR dataset should include embedding coordinates labeled as `emb1` and `emb2`, ensuring one-to-one correspondence with the high-dimensional data through the shared `ID`. The built-in example datasets, `scurve` and `scurve_umap` demonstrates these structures.

To run the entire model pipeline, we can use the `fit_high_model()` function. This function requires: the high-dimensional data (`highd_data`), the embedding data (`nldr_data`), the number of bins along the x-axis (`b1`), the buffer amount as a proportion of the data (`q`), and a benchmark value to identify high-density hexagons (`hd_thresh`).

The function returns an object of class `highd_vis_model` containing the scaled NLDR object (`nldr_scaled_obj`) with three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); the hexagonal object (`hb_obj`), the fitted model in both $2\text{-}D$ (`model_2d`), and $p\text{-}D$ (`model_highd`), and triangular mesh (`trimesh_data`).

```{r, echo=TRUE, eval=TRUE}
model_highd <- fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
```

The resulting model can then be shown in a tour using a two-step process:

```{r, echo=TRUE, eval=TRUE}
combined_data <- comb_data_model(
  highd_data = scurve,
  model_highd = model_highd$model_highd,
  model_2d = model_highd$model_2d
)

tour_view <- show_langevitour(
  point_data = combined_data,
  edge_data = model_highd$trimesh_data
)
```

@@

## Pipeline implementation

@@

In this section, we demonstrate the implementation of each step of the pipeline discussed in the Usage section (`fit_highd_model` and `comb_data_model`). Each step can be run independently to ensure flexibility in the modelling aproach.

@@

### Hexagon binning

The function `calc_bins_y()` determines the configuration of the hexagon grid by computing the number of bins along the y-axis (`b2`), the hexagon binwidth (`a1`), and hexagon binheight (`a2`). This function accepts (i) an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); (ii) the number of bins along the x-axis (`b1`), and (iii) the buffer amount as a proportion (`q`). The buffer ensures that the grid fully covers the data space by extending one hexagon width ($a_1$) and height ($a_2$) beyond the observed data in all directions. By default, $q = 0.1$, but it must be set to a value smaller than the minimum data value to avoid exceeding the data range.

@@

@@

### Obtaining bin centroids

@@

### Indicating neighbors by line segments connecting centroids

To represent the neighborhood structure of hexagonal bins in a $2\text{-}D$ layout, we employ Delaunay triangulation [@lee1980;@albrecht2024] on the centroids of hexagons.

The `tri_bin_centroids()` function generates a triangulation object from the x and y coordinates of hexagon centroids using the `interp::tri.mesh()` function [@albrecht2024].

@@

The `gen_edges()` function uses this triangulation object to extract line segments between neighboring bins. It constructs a unique set of bin-to-bin connections by identifying the triangle edges and filtering duplicate or reversed links. Each edge is then annotated with its start and end coordinates, and a Euclidean distance between the coordinates.

### 2025-11-30 paper-quollr.tex
Move usage to before implementation and re-order many things in implementation

@@

@@

The following demonstration of the package's functionality assumes \texttt{quollr} has been loaded. To begin the workflow, user needs two inputs: the high-dimensional dataset and the corresponding NLDR data. The high-dimensional data must contain a unique \texttt{ID} column, with data columns prefixed by the letter \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, etc.). The NLDR dataset should include embedding coordinates labeled as \texttt{emb1} and \texttt{emb2}, ensuring one-to-one correspondence with the high-dimensional data through the shared \texttt{ID}.

To illustrate the workflow, we use the built-in example dataset \texttt{scurve}, a \(7\text{-}D\) simulated dataset consisting of \(1000\) observations. The first three variables define a \(3\text{-}D\) S-shaped manifold, while the remaining four variables introduce low-magnitude uniform noise, yielding a structured yet noisy high-dimensional dataset.

\begin{verbatim}
data("scurve")
\end{verbatim}

For this example, we use the UMAP layout, which is used \texttt{n\_neighbors\ =} \(46\) and \texttt{min\_dist\ =} \(0.9\) as hyper-parameters.

\begin{verbatim}
library(umap)
library(dplyr)

scurve_umap <- umap(
  scurve |> select(-ID),
  config = modifyList(umap.defaults, list(
    n_neighbors = 46,
    n_components = 2,
    min_dist = 0.9
  ))
)$layout |>
  as_tibble(.name_repair = ~ paste0("emb", 1:2)) |>
  mutate(ID = scurve$ID)
\end{verbatim}

\subsection{Main function}\label{main-function}

The mains steps for the algorithm can be executed by the main function \texttt{fit\_highd\_model()}, or can be run separately for more flexibility.

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{hd\_thresh}). The function returns an object of class \texttt{highd\_vis\_model} containing the scaled NLDR object (\texttt{nldr\_scaled\_obj}) with three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); the hexagonal object (\texttt{hb\_obj}), the fitted model in both \(2\text{-}D\) (\texttt{model\_2d}), and \(p\text{-}D\) (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

\begin{verbatim}
fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
\end{verbatim}

>>>>

The user needs two inputs: the high-dimensional dataset and the corresponding NLDR data. The high-dimensional data must contain a unique \texttt{ID} column, with data columns prefixed by the letter \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, etc.). The NLDR dataset should include embedding coordinates labeled as \texttt{emb1} and \texttt{emb2}, ensuring one-to-one correspondence with the high-dimensional data through the shared \texttt{ID}. The built-in example datasets, \texttt{scurve} and \texttt{scurve\_umap} demonstrates these structures.

To run the entire model pipeline, we can use the \texttt{fit\_high\_model()} function. This function requires: the high-dimensional data (\texttt{highd\_data}), the embedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of the data (\texttt{q}), and a benchmark value to identify high-density hexagons (\texttt{hd\_thresh}).

The function returns an object of class \texttt{highd\_vis\_model} containing the scaled NLDR object (\texttt{nldr\_scaled\_obj}) with three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); the hexagonal object (\texttt{hb\_obj}), the fitted model in both \(2\text{-}D\) (\texttt{model\_2d}), and \(p\text{-}D\) (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

\begin{verbatim}
model_highd <- fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
\end{verbatim}

The resulting model can then be shown in a tour using a two-step process:

\begin{verbatim}
combined_data <- comb_data_model(
  highd_data = scurve,
  model_highd = model_highd$model_highd,
  model_2d = model_highd$model_2d
)

tour_view <- show_langevitour(
  point_data = combined_data,
  edge_data = model_highd$trimesh_data
)
\end{verbatim}

@@

\subsection{Pipeline implementation}\label{pipeline-implementation}

@@

In this section, we demonstrate the implementation of each step of the pipeline discussed in the Usage section (\texttt{fit\_highd\_model} and \texttt{comb\_data\_model}). Each step can be run independently to ensure flexibility in the modelling aproach.
